Supplementary Material: Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-09-21 manuscript/main.qmd
seasonality paragraph, updating results, discussion, SM

```{=html}
<!-- discarded titles:
Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene
A Bayesian model of latitudinal temperature gradients applied to the Early Eocene
Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene
Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient
Combining geochemical and ecological proxies...
Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
"An Early Eocene temperature gradient modelled with geological and ecological proxies"
-->
```
Accurately reconstructing large-scale palaeoclimate patterns from sparse local records is critical for understanding the evolution of Earth's climate. Particular challenges arise from the patchiness, uneven spatial distribution, and disparate nature of palaeoclimatic proxy records. Geochemical data typically provide temperature estimates via transfer functions derived from experiments. Similarly, transfer functions based on the climatic requirements of modern taxa exist for some fossil groups, such as pollen assemblages. In contrast, most ecological and lithological data (e.g. coral reefs and evaporites) only convey information on broad climatic requirements. Historically, most large-scale proxy-based reconstructions have used either geochemical or ecological data, but few studies have combined multiple proxy types into a single quantitative reconstruction. Large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into account the spatial distribution of samples, leading to biased temperature reconstructions. Here, we present a Bayesian hierarchical model to integrate ecological data with established geochemical proxies into a unified quantitative framework, bridging gaps in the latitudinal coverage of proxy data. We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal sea surface temperature gradient and global average temperatures of the EECO. Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data. We show that integrating multiple types of proxy data, and adequate prior information, has the potential to substantially reduce uncertainty in palaeoclimate reconstructions, allowing for unbiased temperature estimates from sparse data.

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (latitudes of EECO samples). Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme icehouse, icehouse, greenhouse, and extreme greenhouse. The black line illustrates the simulated gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1,000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column). The blue shadings depict the 90%, 95%, and 99% credible intervals. Bold black text within each panel depicts the coefficient of determination ($R^2$) for estimating goodness of fit between the simulated and modelled gradient. The median (50%) $R^2$ value along with the 95% credible intervals from all model runs are shown. Each gradient is depicted in absolute latitude.
:::

@@

The Eocene temperature gradient reconstructed with our Bayesian model is starkly different from the modern (Fig 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` -- `r tdiff_lat0$"97.5%"`)&deg;C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` -- `r tdiff_lat90$"97.5%"`)&deg;C higher. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` -- `r eocene_gradient$"97.5%"`)&deg;C for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern. To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is slightly lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` -- `r eocene_gradient_polar_circle$"97.5%"`)&deg;C.

The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient -- `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` -- `r eeco_sdy$"97.5%"`)&deg;C -- which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`. This signifies that the early Eocene data does not fit as well to the logistic latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` -- `r eotemp_global$"97.5%"`)&deg;C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern. A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` -- `r eoceneproxy_global_diff$"97.5%"`)&deg;C. The median latitudinal gradient is higher near the equator and in high latitudes when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`&deg;C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Fig. S2). This indicates that the ecological proxy data provide additional constraints on the shape of the early Eocene temperature gradient. In contrast, including ecological proxies, but widening the uncertainty around the low-latitude ecological proxy data does not significantly change the resulting gradient (Fig. S4).

@@

Our results show that our Bayesian model can be used to reconstruct different types of latitudinal SST gradients from proxy data, even with small sample sizes (*n* = 10 -- 20) and patchy sampling distributions (Fig. 2). This is an advancement over previously used linear, quadratic or Gaussian approximations [e.g. @Bijl2009; @Tierney2017], which can fit only specific types of gradients. As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients, which are sometimes favoured as they can flexibly follow the shape of an unknown temperature gradient [e.g. @Zhang2019; @Jones2022]. However, when used for interpolation or prediction outside the proxy range, non-parametric methods such as Gaussian process regression strictly respond to the data [e.g. @Inglis2020]. This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty, dictate the reconstruction of large-scale climate patterns, without the option of including additional knowledge (e.g. that latitudinal temperature gradients should be broadly negative).

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020]. Our polar circle to equatorial gradient estimate is lower than most previous estimates at 7.8&deg;C, although the 95% credible interval extends up to 13.7&deg;C and thus overlaps earlier estimates based on shallow water proxies. The confirmation of a very flat gradient by both geochemical and ecological shallow water data indicates that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.

@@

Discrepancies between earlier, proxy-based reconstructions and our modelling results are most pronounced in latitudes beyond the polar circle, as earlier approaches [e.g. @Tierney2017] predict almost linearly decreasing SSTs towards the poles, whereas our median prediction suggests only a slight decrease beyond the polar circle. The scarcity of temperature records in this range leads to widening credible intervals in our prediction, including the possibility of stronger temperature decreases. Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable. However, the presence of high proxy-derived temperature estimates at \~ 60&deg; latitudes forces the modelled median temperature curve to be too high at \~ 24&deg;C, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5&deg;C). In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@Pross2012; @Popescu2021].

The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMST. Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019]. Despite excluding &delta;^18^O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX~86~, and cooler temperatures derived from &delta;^18^O, &Delta;~47~, and the ecological proxies. Seasonality [@Keating-Bitonti2011] and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.

>>>>

```{=html}
<!-- discarded titles:
Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene
A Bayesian model of latitudinal temperature gradients applied to the Early Eocene
Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene
Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient
Combining geochemical and ecological proxies...
Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
"An Early Eocene temperature gradient modelled with geological and ecological proxies"
-->
```
Accurately reconstructing large-scale palaeoclimate patterns from sparse local records is critical for understanding the evolution of Earth's climate. Particular challenges arise from the patchiness, uneven spatial distribution, and disparate nature of palaeoclimatic proxy records. Geochemical data typically provide temperature estimates via transfer functions derived from experiments. Similarly, transfer functions based on the climatic requirements of modern taxa exist for some fossil groups, such as pollen assemblages. In contrast, most ecological and lithological data (e.g. coral reefs and evaporites) only convey information on broad climatic requirements. Historically, most large-scale proxy-based reconstructions have used either geochemical or ecological data, but few studies have combined multiple proxy types into a single quantitative reconstruction. Large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into account the spatial distribution of samples, leading to biased temperature reconstructions. Here, we present a Bayesian hierarchical model to integrate ecological data with established geochemical proxies into a unified quantitative framework, bridging gaps in the latitudinal coverage of proxy data. We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal sea surface temperature gradient and global average temperatures of the EECO. Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data. We show that integrating multiple types of proxy data, and adequate prior information, has the potential to enhance quantiative palaeoclimate reconstructions, allowing for more accurate temperature estimates from sparse data.

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (latitudes of EECO samples). Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme icehouse, icehouse, greenhouse, and extreme greenhouse. The black line illustrates the simulated gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1,000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column). To generate the random samples, noise from a normal distribution with a standard deviation of 3.8 was added to all temperatures. The blue shadings depict the 90%, 95%, and 99% credible intervals. Bold black text within each panel depicts the coefficient of determination ($R^2$) for estimating goodness of fit between the simulated and modelled gradient. The median (50%) $R^2$ value along with the 95% credible intervals from all model runs are shown. Each gradient is depicted in absolute latitude.
:::

@@

The Eocene temperature gradient reconstructed with our Bayesian model is starkly different from the modern (Fig. 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` -- `r tdiff_lat0$"97.5%"`)&deg;C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` -- `r tdiff_lat90$"97.5%"`)&deg;C higher. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` -- `r eocene_gradient$"97.5%"`)&deg;C for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern. To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is slightly lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` -- `r eocene_gradient_polar_circle$"97.5%"`)&deg;C.

The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient -- `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` -- `r eeco_sdy$"97.5%"`)&deg;C -- which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`. This is illustrated by the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` -- `r eotemp_global$"97.5%"`)&deg;C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern. A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` -- `r eoceneproxy_global_diff$"97.5%"`)&deg;C. The median modelled temperature is higher near the equator and in high latitudes when excluding the ecological proxies, with a flattened median gradient of `r eocene_proxy_gradient$"50%"`&deg;C (Fig. S2). In contrast, including ecological proxies, but widening the uncertainty around the low-latitude ecological proxy data does not significantly change the resulting gradient (Fig. S3).

@@

Our results show that our Bayesian model can be used to reconstruct different types of latitudinal SST gradients from proxy data with moderate sample sizes (*n* = 10 -- 34) and patchy sampling distributions (Fig. 2). This is an advancement over previously used linear, quadratic or Gaussian approximations [e.g. @Bijl2009; @Tierney2017], which can fit only specific types of gradients. As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients, which are sometimes favoured as they can flexibly follow the shape of an unknown temperature gradient [e.g. @Zhang2019; @Jones2022]. However, when used for interpolation or prediction outside the proxy range, non-parametric methods such as Gaussian process regression strictly respond to the data [e.g. @Inglis2020]. This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty, dictate the reconstruction of large-scale climate patterns, without the option of including additional knowledge (e.g. that latitudinal temperature gradients should be broadly negative).

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020]. Our median polar circle to equatorial gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C. The confirmation of a very flat gradient by both geochemical and ecological shallow water data indicates that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.

@@

<!-- OUTDATED Discrepancies between earlier, proxy-based reconstructions and our modelling results are most pronounced in latitudes beyond the polar circle, as earlier approaches [e.g. @Tierney2017] predict almost linearly decreasing SSTs towards the poles, whereas our median prediction suggests only a slight decrease beyond the polar circle. The scarcity of temperature records in this range leads to widening credible intervals in our prediction, including the possibility of stronger temperature decreases. Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable. However, the presence of high proxy-derived temperature estimates at \~ 60&deg; latitudes forces the modelled median temperature curve to be too high at \~ 24&deg;C, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5&deg;C). In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@Pross2012; @Popescu2021]. -->

The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMST. Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019]. Despite excluding &delta;^18^O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX~86~, and cooler temperatures derived from &delta;^18^O, &Delta;~47~, and the ecological proxies.

Temporal changes within the EECO [@Westerhold2018], and seasonality [@Keating-Bitonti2011; @Ivany2022] may also contribute to the large variability of the EECO proxy data. Based on the occurrence of heterotrophic carbonates, @Davies2019 suggest that mid- and high-latitude geochemical proxy data from the EECO may be biased towards summer temperatures. Some of the geochemical mid-latitude geochemical proxy data from @Hollis2019 may therefore suggest higher than actual mean annual temperatures, and the variability of temperature estimates from individual localities is higher in mid - high latitudes (Fig. S6). Critically, however, the mangrove data strongly supports our inference of a flattened gradient independent of the geochemical proxy record.

### 2023-09-21 manuscript/supplementary.qmd
seasonality paragraph, updating results, discussion, SM

Estimated latitudinal temperature gradient using only wider ecological limits for the coral data and for the *Avicennia*-Rhizophoraceae mangrove assemblage (orange), showing the median (line) and 95% credible interval (shading). The ecological limits were expanded to include $95\%$ of the probability density in the interval of $16.0$ to $35.6$&deg;C, using the minimal monthly temperature experienced by coral reefs (@Kleypas1999), and a theoretical maximum derived by [@jones2022reef]. Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$. The black line and shading in the background show the latitudinal temperature gradient with the narrower ecological limits, as in Figure 4.
:::

@@

# EECO gradient with uncertainty on proxy observations 
Most of the geochemical proxy data used in the EECO analysis (300 of 308) come with uncertainties around their temperature estimates. Due to their negligible influence on the model results, we have not included them in the main analysis, but we show a comparison between the main EECO model output (see Fig. 4) and an the results of an expanded model, including uncertainties, below.

@@

<!-- REMOVED
```{r out.width='100%', echo = FALSE, fig.cap = "Fig S4. Simulated latitudinal temperature gradients sampled at the absolute latitudes of the EECO data. The dark grey points show the simulated temperature data, the black line shows the shape of the true gradient, and the medians (lines) and 95% credible intervals (shadings) of the inferred gradients using the model are shown in red. \\\n a) linear gradient; $SST = 30-|latitude|/3$ \\\n b) flat gradient; $SST = 25$ \\\n c) quadratic gradient; $SST = 30-|latitude|^2/270$ \\\n d) as c), but the priors on A, M, and B were relaxed to allow for a better fit to the quadratic gradient in high latitudes \\\n  All latitudinal temperature gradients are well captured by the model in the latitudinal range for which data is provided. If the priors, particularly the prior on the lower limit of the gradient (A), are relaxed (d), the model can approximate a quadratic or linear function even in high latitudes and at cold temperatures."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS4_different_gradients.png", error = FALSE)
```
-->

>>>>

Estimated latitudinal temperature gradient using only wider ecological limits for the coral data and for the *Avicennia*-Rhizophoraceae mangrove assemblage (orange), showing the median (line) and 95% credible interval (shading). The ecological limits were expanded to include $95\%$ of the probability density in the interval of $16.0$ to $35.6$&deg;C, using the minimal monthly temperature experienced by coral reefs (Kleypas et al., 1999), and a theoretical maximum derived by Jones et al. (2022). Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$. The black line and shading in the background show the latitudinal temperature gradient with the narrower ecological limits, as in Figure 4.
:::

@@

## EECO gradient with uncertainty on proxy observations 
Most of the geochemical proxy data used in the EECO analysis (300 of 308) come with uncertainties around their temperature estimates. Due to their negligible influence on the model results, we have not included them in the main analysis, but we show a comparison between the main EECO model output (see Fig. 4) and an the results of an expanded model, including uncertainties, below.

@@

::: {#fig-S6}
![](../figures/SM/FigS_temp_location_sd.png){width="8.5cm"}

Standard deviation of temperature estimates from individual localities of the EECO geochemical proxy data from Hollis et al. (2019) with more than one sample, against the absolute latitude of the locality (n = 17). The line shows a linear regression, suggesting a moderate effect of absolute latitude on proxy data variability, with mid- and hight-latitude data being more variable than low-latitude data.
:::
<!-- REMOVED
```{r out.width='100%', echo = FALSE, fig.cap = "Fig S4. Simulated latitudinal temperature gradients sampled at the absolute latitudes of the EECO data. The dark grey points show the simulated temperature data, the black line shows the shape of the true gradient, and the medians (lines) and 95% credible intervals (shadings) of the inferred gradients using the model are shown in red. \\\n a) linear gradient; $SST = 30-|latitude|/3$ \\\n b) flat gradient; $SST = 25$ \\\n c) quadratic gradient; $SST = 30-|latitude|^2/270$ \\\n d) as c), but the priors on A, M, and B were relaxed to allow for a better fit to the quadratic gradient in high latitudes \\\n  All latitudinal temperature gradients are well captured by the model in the latitudinal range for which data is provided. If the priors, particularly the prior on the lower limit of the gradient (A), are relaxed (d), the model can approximate a quadratic or linear function even in high latitudes and at cold temperatures."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS4_different_gradients.png", error = FALSE)
```
-->
